### 2024-03-29 doc/15yrsChange.Rmd
updating rmarkdown citations

---
title: ''
output: 
    #pdf_document: 
    bookdown::word_document2: 
      md_extensions: +raw_html-markdown_in_html_blocks
      pandoc_args: ['--lua-filter', 'read_html.lua']
      reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
  #officedown::rdocx_document
bibliography: references.bib
csl: new-phytologist.csl

@@

The temperate forest biome plays a critical role in the global carbon cycle, accounting for almost half of the net global forest carbon (C) sink [@harris_global_2021], with temperate deciduous forests alone sequestering >300 Tg C yr^-1^ [@pugh_role_2019]. Although currently a net C sink, the future of the temperate forest biome remains uncertain.
The dominant pattern projected by Earth System Models is a continued, albeit weakening, temperate forest C sink for the remainder of the 21st century [@wu_uncertainty_2023; @ahlström_robustness_2012]. 
Yet, global C models predict a wide range of future trajectories of CO~2~ sequestration [@ahlström_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014; @clark_continentwide_2021].
One mechanism that is not represented in global C models is the impact of nuisance species – i.e., indigenous or non-indigenous species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, resulting in undesirable ecological consequences [@skandrani_nuisance_2014; nunez_population_2010]. 
Similarly, the impact of nuisance species on forest carbon budgets is not considered in machine learning or niche models that seek to project future forest distribution and carbon stocks, nor in carbon offset projects [@wu_uncertainty_2023].
This is problematic in that nuisance species – including non-indigenous insect pests and pathogens, non-indigenous plants, and over-abundant herbivores – are dramatically impacting carbon cycling in temperate deciduous forests around the world.

Non-indigenous pests and pathogens can have a protracted impact on forests, often modifying forest productivity, nutrient cycling, successional trajectories, and geomorphic processes [@lovett_nonnative_2016; @fischer_disturbances_2013; @herms_emerald_2014; @fei_biogeomorphic_2014]. Historically introduced to new ecosystems through intercontinental trade, non-indigenous forest insects and diseases will continue to proliferate with increased globalization and human movement [@fischer_disturbances_2013; @aukema_historical_2010; @brockerhoff_ecology_2006]. Many of the prominent pests and pathogens in temperate deciduous forests have dramatic effects on canopy composition and structure, decimating once-abundant canopy species [@aukema_historical_2010; @anderson-teixeira_longterm_2021a; @klooster_ash_2014; @ellison_present_2018]. Since many pests and pathogens do not currently occupy the full extent of their host species range, a significant portion of the remaining forest canopy is susceptible to future invasion from nuisance species [@fei_biomass_2019], further increasing potentials for biomass loss and altered carbon cycle dynamics.

Another category of nuisance species is mammals whose population densities have been inflated due to human activities [@murphy_invasive_2014; @clout_invasion_2008], including white-tailed deer (*Odocoileus virginianus*) in U.S. eastern deciduous forests. Their populations have dramatically increased in the past 50 years (McShea et al. 1997), due in large part to global change (McGarvey et al. 2013, Brown et al. 2000, Côté et al. 2004, Rooney 2001). White-tailed deer preferentially browse on woody plant species in the seedling and sapling life stages, which reduces survival, growth and density [@mcgarvey_effects_2013a]. Chronic white-tailed deer overabundance can reduce understory plant diversity and decrease the abundance of traditionally dominant species [@bugalho_effects_2013a] while giving an advantage to species they find less palatable, including both indigenous (e.g., pawpaw, *Asimina tribola*; spicebush, *Lindera benzoin*) and non-indigenous (e.g., XX ) species (REFS). The strong impact of deer browsing, exacerbated by competition with dense deer-resistant understory vegetation and possibly climate change  results in dramatic reduction of canopy tree species in the understory [@mcgarvey_effects_2013a; @russell_interactions_2017; @miller_compounding_2019; @gorchov_differential_2021; @miller_overabundant_2023].

@@

In addition to our consideration of the entire forest plot, we focus further on three portions of the plot, all upland forest (i.e., excluding low-lying areas around streams), with varying levels of exposure non-indigenous insects and pathogens and white-tailed deer (Fig. 1).  The first portion we define as“Low deer, low canopy vulnerability”: A 4-ha portion of upland forest in the SE quarter of the plot (Fig. 1a), fenced in 1990 and maintained with only occasional deer presence over the past 24 years. This portion has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens [@holm_interactive_2013]. The second focal area is a “high deer, low canopy vulnerability” 16.3 ha upland forest area outside the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens (Fig. 1b). The third focal portion of the plot has high deer, high canopy vulnerability: corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens. We defined upland forest according to topographic wetness index, which was calculated for the forest plot by mcgregor_tree_2021. We defined vulnerability to non-indigenous insects and pathogens based on the abundance of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment (2008). The following canopy species were identified as vulnerable: ash (Fraxinus spp.), elm (Ulmus spp.), butternut (Juglans cinerea), and American chestnut (Carya dentata) [@anderson-teixeira_longterm_2021a]

@@

As part of the ForestGEO network, the plot undergoes a comprehensive woody plant inventory every five years according to the protocol detailed in Condit (1998). This inventory, hereafter referred to as the census, includes all stems greater than one centimeter in diameter at breast height (1.3 meters; dbh). The census records the dbh, species, living status, and spatial location of each censused stem. Each individual is marked with an identifying number for sequential data collection in subsequent censuses. In multi-stemmed individuals, each additional stem that surpasses the dbh threshold receives a stem number and associated tag. The plot was established in 2008 and recensused three times (2013, 2018, 2023) since, comprising 15 years of forest dynamics data. In the last census, data cleaning was continuously implemented 1) in an ESRI ArcGIS Field Maps application via built-in checks and 2) on a daily basis, via error reports generated through GitHub actions (Kim et al., 2022). Each individual’s location was recorded in the Field Maps application using a digitized map of the plot (citation). In addition to the ForestGEO census, a tree health and mortality survey was conducted annually starting in 2014 on all trees with a dbh ≥ 10 cm in the previous census (Gonzalez-Akre et al. 2016). Data were collected on the current status, canopy position, percentage of crown living and intact, and visible indicators of tree health, such as physical damage, potential pathogens and insect infestation.

@@

We estimated total aboveground biomass (AGB) from the dbh of each stem in each census using extra-tropical allometric equations from the R package *allodb* (Gonzalez-Akre, 2024). The *allodb* package creates species-specific allometric equations by weighting all applicable allometries within its database filtered by tree size, geography, and taxonomic information. Using the calculated stem AGB, we estimated AGB stocks at each census year for each of the plot study regions and the entire 25.6-ha plot. We quantified trends in AGB over time by calculating annualized AGB fluxes including: 1) the net change in AGB, 2) aboveground woody growth (AWG; the addition of woody biomass from tree growth), 3) aboveground woody mortality (AWM; the change of living biomass to necromass due to tree mortality), and 4) aboveground woody recruitment (AWR; the addition of living biomass through recruitment) for each census interval and plot region. We further assessed the role of nuisance species on mortality by calculating AWM flux for each year and each species using the annual mortality survey data.

>>>>

---
title: ''
output: 
    #pdf_document: 
    bookdown::word_document2: 
      md_extensions: +raw_html-markdown_in_html_blocks
      pandoc_args: ['--lua-filter', 'read_html.lua']
      reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
  #officedown::rdocx_document
bibliography: citations.bib
csl: new-phytologist.csl

@@

The temperate forest biome plays a critical role in the global carbon cycle, accounting for almost half of the net global forest carbon (C) sink [@harris_global_2021], with temperate deciduous forests alone sequestering >300 Tg C yr^-1^ [@pugh_role_2019]. Although currently a net C sink, the future of the temperate forest biome remains uncertain.
The dominant pattern projected by Earth System Models is a continued, albeit weakening, temperate forest C sink for the remainder of the 21st century [@wu_uncertainty_2023; @ahlstrom_robustness_2012]. 
Yet, global C models predict a wide range of future trajectories of CO~2~ sequestration [@ahlstrom_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014; @clark_continentwide_2021].
One mechanism that is not represented in global C models is the impact of nuisance species – i.e., indigenous or non-indigenous species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, resulting in undesirable ecological consequences [@skandrani_nuisance_2014; @nunez_population_2010]. 
Similarly, the impact of nuisance species on forest carbon budgets is not considered in machine learning or niche models that seek to project future forest distribution and carbon stocks, nor in carbon offset projects [@wu_uncertainty_2023].
This is problematic in that nuisance species – including non-indigenous insect pests and pathogens, non-indigenous plants, and over-abundant herbivores – are dramatically impacting carbon cycling in temperate deciduous forests around the world.

Non-indigenous pests and pathogens can have a protracted impact on forests, often modifying forest productivity, nutrient cycling, and successional trajectories. [@lovett_nonnative_2016; @herms_emerald_2014]. Historically introduced to new ecosystems through intercontinental trade, non-indigenous forest insects and diseases will continue to proliferate with increased globalization and human movement [@fischer_disturbances_2013; @aukema_historical_2010; @brockerhoff_ecology_2006]. Many of the prominent pests and pathogens in temperate deciduous forests have dramatic effects on canopy composition and structure, decimating once-abundant canopy species [@aukema_historical_2010; @anderson-teixeira_longterm_2021; @klooster_ash_2014; @ellison_present_2018]. Since many pests and pathogens do not currently occupy the full extent of their host species range, a significant portion of the remaining forest canopy is susceptible to future invasion from nuisance species [@fei_biomass_2019], further increasing potentials for biomass loss and altered carbon cycle dynamics.

Another category of nuisance species is mammals whose population densities have been inflated due to human activities [@murphy_invasive_2014; @clout_invasion_2008], including white-tailed deer (*Odocoileus virginianus*) in U.S. eastern deciduous forests. Their populations have dramatically increased in the past 50 years [@mcshea_science_2003], due in large part to global change [cote_ecological_2004; @rooney_deer_2001; @brown_future_2000]. White-tailed deer preferentially browse on woody plant species in the seedling and sapling life stages, which reduces survival, growth and density [@mcgarvey_effects_2013a]. Chronic white-tailed deer overabundance can reduce understory plant diversity and decrease the abundance of traditionally dominant species [@bugalho_effects_2013a] while giving an advantage to species they find less palatable, including both indigenous (e.g., pawpaw, *Asimina tribola*; spicebush, *Lindera benzoin*) and non-indigenous (e.g., XX ) species (REFS). The strong impact of deer browsing, exacerbated by competition with dense deer-resistant understory vegetation and possibly climate change  results in dramatic reduction of canopy tree species in the understory [@mcgarvey_effects_2013a; @russell_interactions_2017; @miller_compounding_2019; @gorchov_differential_2021; @miller_overabundant_2023].

@@

In addition to our consideration of the entire forest plot, we focus further on three portions of the plot, all upland forest (i.e., excluding low-lying areas around streams), with varying levels of exposure non-indigenous insects and pathogens and white-tailed deer (Fig. 1).  The first portion we define as“Low deer, low canopy vulnerability”: A 4-ha portion of upland forest in the SE quarter of the plot (Fig. 1a), fenced in 1990 and maintained with only occasional deer presence over the past 24 years. This portion has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens [@holm_interactive_2013]. The second focal area is a “high deer, low canopy vulnerability” 16.3 ha upland forest area outside the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens (Fig. 1b). The third focal portion of the plot has high deer, high canopy vulnerability: corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens. We defined upland forest according to topographic wetness index, which was calculated for the forest plot by mcgregor_tree_2021. We defined vulnerability to non-indigenous insects and pathogens based on the abundance of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment (2008). The following canopy species were identified as vulnerable: ash (Fraxinus spp.), elm (Ulmus spp.), butternut (Juglans cinerea), and American chestnut (Carya dentata) [@anderson-teixeira_longterm_2021]

@@

As part of the ForestGEO network, the plot undergoes a comprehensive woody plant inventory every five years according to the protocol detailed in @condit_tropical_1998. This inventory, hereafter referred to as the census, includes all stems greater than one centimeter in diameter at breast height (1.3 meters; dbh). The census records the dbh, species, living status, and spatial location of each censused stem. Each individual is marked with an identifying number for sequential data collection in subsequent censuses. In multi-stemmed individuals, each additional stem that surpasses the dbh threshold receives a stem number and associated tag. The plot was established in 2008 and recensused three times (2013, 2018, 2023) since, comprising 15 years of forest dynamics data. In the last census, data cleaning was continuously implemented 1) in an ESRI ArcGIS Field Maps application via built-in checks and 2) on a daily basis, via error reports generated through GitHub actions [@kim_implementing_2022]. Each individual’s location was recorded in the Field Maps application using a digitized map of the plot (citation). In addition to the ForestGEO census, a tree health and mortality survey was conducted annually starting in 2014 on all trees with a dbh ≥ 10 cm in the previous census [@gonzalez-akre_patterns_2016]. Data were collected on the current status, canopy position, percentage of crown living and intact, and visible indicators of tree health, such as physical damage, potential pathogens and insect infestation.

@@

We estimated total aboveground biomass (AGB) from the dbh of each stem in each census using extra-tropical allometric equations from the R package *allodb* [@gonzalez-akre_allodb_2022]. The *allodb* package creates species-specific allometric equations by weighting all applicable allometries within its database filtered by tree size, geography, and taxonomic information. Using the calculated stem AGB, we estimated AGB stocks at each census year for each of the plot study regions and the entire 25.6-ha plot. We quantified trends in AGB over time by calculating annualized AGB fluxes including: 1) the net change in AGB, 2) aboveground woody growth (AWG; the addition of woody biomass from tree growth), 3) aboveground woody mortality (AWM; the change of living biomass to necromass due to tree mortality), and 4) aboveground woody recruitment (AWR; the addition of living biomass through recruitment) for each census interval and plot region. We further assessed the role of nuisance species on mortality by calculating AWM flux for each year and each species using the annual mortality survey data.
